Our analysis offers insights into how renewable energy imports might reduce
overall systems costs and interact with European energy infrastructure. Our
results show that imports of green energy reduce the costs of a carbon-neutral
European energy system by \bneuro{37} (4.4\%), noting, however, that the
uncertainty range is considerable. While we find that some imports are
beneficial within a $\pm$20\% variation of import costs, system cost savings
range between 1\% and 10\%. However, what is consistent within this range are
the diminishing returns of energy imports for larger quantities, with peak cost
savings below imports of 3000~TWh/a (equivalent to 90~Mt of hydrogen). We also find
that there is value in pursuing some \mbox{power-to-X} production in Europe as a
source of flexibility for wind and solar integration and as a potential source
of waste heat for district heating networks. Another siting factor favouring
European \mbox{power-to-X} is the wide availability of sustainable biogenic and
industrial carbon sources, which helps reduce reliance on more costly direct air
capture.

\todo[inline]{comparison to genge for 2030 and 2050 costs}

In relation to other studies, our assumed import costs for different carriers
mostly conform to the interquartile ranges for 2050 of the meta-study by Genge
et al.~\cite{gengeSupplyCostsGreen2023} for imports to Europe. Some higher
costs, e.g.~for ammonia, can be attributed to our updated electrolyser cost
assumptions (950\euro{}~kW$_\text{el.}^{-1}$ in 2040), which reflect recent
market developments.\cite{ieaGlobalHydrogenReview2024} Also for steel imports,
our central cost estimate of 531~\euro{}~t$^{-1}$ for the lowest-cost exporter
is positioned between studies with lower\cite{lopezDefossilisedSteel2023} and
higher\cite{verpoortImpactGlobalHeterogeneity2024} cost estimates. Among other
studies investigating the relationship between energy imports and the European
energy system, several analyses report lower import shares in the range of
10-20\% of total hydrogen
supply.\cite{seckHydrogenDecarbonization2022,frischmuthHydrogenSourcingStrategies2022,kountourisUnifiedEuropeanHydrogen2024}
For instance, Kountouris et al.~\cite{kountourisUnifiedEuropeanHydrogen2024} see
limited hydrogen imports of 182~TWh~a$^{-1}$ from the Maghreb region and Ukraine
despite favourable import costs of 33~\euro{}~MWh$_{H_2}^{-1}$. Conversely,
Wetzel et al.~\cite{wetzelGreenEnergy2023a} find higher import shares of 53\%
for methane and 43\% for hydrogen. The latter closely aligns with our 49\%
import share for hydrogen. Wetzel et al.~\cite{wetzelGreenEnergy2023a} find that
imports reduce system costs by 2.8\%, which is also comparable to our 2.4\%
system cost reduction when only direct hydrogen imports are considered. The most
pronounced import dependency we found in Schmitz et
al.\cite{schmitzImplicationsHydrogenImport2024a}, with import shares beyond 90\%
for Germany. Results in Kountouris et
al.~\cite{kountourisUnifiedEuropeanHydrogen2024} further substantiate our
finding that derivative imports and demand relocation could diminish hydrogen
network benefits.

Several limitations of our study should be noted. First, the optimization
results represent a long-term equilibrium that disregards potential
transition-related infrastructure lock-ins or mid-term ramp-up constraints of
export capacities or domestic infrastructure development. A further limitation
is that our cost-based analysis of imports, which best reflects long-term
bilateral purchase agreements, neglects price impacts of intensifying global
competition for green fuel imports and exports.\cite{galimovaGlobalTrading2023a}
Besides unclear market developments, local challenges in exporting regions such
as public acceptance for export-oriented energy
projects\cite{ishmamMappingLocalGreen2024} and potential water
scarcity\cite{franzmannGreenHydrogenCostpotentials2023,terlouwFutureHydrogenEconomies2024}
to produce large amounts of hydrogen in renewable-rich but arid regions are not
addressed. Furthermore, the model's lack of spatial resolution for CO$_2$ means
that carbonaceous hydrogen derivatives are sited where H$_2$ is cheapest,
implicitly assuming that the CO$_2$ from biogenic or industrial sources can be
transported there. However, such required CO$_2$ pipelines could be built at
relatively low additional system cost.\cite{hofmannH2CO2Network2024} In the
context of carbon management, more lenient assumptions on sustainable biofuel
potentials, allowed levels of geological carbon sequestration, or plastic
landfill could alter the results, shifting the system away from synthetic
electrofuels towards more fossil fuel use with carbon capture or carbon dioxide
removal.\cite{hofmannH2CO2Network2024,millingerDiversityBiomassUsage2023}

\todo[inline]{uncertain technological development}

Overall, we find that the import vectors used strongly affect domestic
infrastructure needs. For example, only a smaller hydrogen network would be
required if hydrogen derivatives were largely imported and the domestic ammonia
and steel industry is allowed to relocate. We also identify higher electricity
backup requirements in the absence of large power-to-X flexibilities. These
findings underscore the importance of coordination between energy import
strategies and infrastructure policy decisions. Our results present a
quantitative basis for further discussions about the trade-offs between system
cost, carbon neutrality, public acceptance, energy security, infrastructure
buildout, and imports.

The small differences in cost observed between some scenarios are particularly
relevant because factors other than pure costs might then drive the designs of
import strategies. The relatively limited cost benefit of imports and offshoring
of industrial production may speak against imports. Concerns about energy
sovereignty would motivate more domestic supply and diversified imports. For
instance, shipborne imports of hydrogen derivatives would reduce pipeline
lock-in and mitigate the risks of sudden supply disruptions and abuse of market
power. From a practical perspective, it may also be more appealing to focus on
carriers that are already globally traded commodities and to prefer
infrastructure offering quick deployment.

Policymakers in Europe might prefer such easy-to-implement systems featuring,
for instance, lower domestic infrastructure requirements, reuse of existing
infrastructure, lower technology risk, and reduced land usage for broader public
support than the most cost-effective solution. Moreover, policies favoring local
energy supply chains and importing intermediary products like sponge iron could
preserve European jobs while outsourcing only the most energy-intensive
processes. However, in shifting potential land use and infrastructure conflicts
to abroad, where population densities are often lower, potential exporting
countries must weigh the prospect of economic development against internal
social and environmental concerns, particularly in countries with a history of
colonial exploitation.\cite{tunnGreenHydrogenTransitions2024} Ultimately,
Europe's energy strategy must balance cost savings from green energy and
material imports with broader concerns like geopolitics, economic development,
public opinion, and the willingness of potential exporting countries in order to
ensure a swift, secure, and sustainable energy future. Our research shows that
there is maneuvering space to accommodate such non-cost concerns.
