

\subsection*{PyPSA-Eur: Overview of European energy system model}
\label{sec:methods-overview}

For our analysis, we use the European sector-coupled high-resolution energy
system model PyPSA-Eur\cite{horschPyPSAEurOpen2018a} (derivative of v0.13.0) based on the open-source
modelling framework PyPSA\cite{brownPyPSAPython2018} (Python for Power System
Analysis), covering the energy demands of all sectors including electricity,
heat, transport, industry, agriculture, as well as non-energy feedstock demands,
international shipping, and aviation. An overview of considered supply,
consumption, and balancing technologies per carrier is shown in
\sfigref{fig:si:supply-consumption-options}.

The model simultaneously optimises spatially explicit investments and the
operation of generation, storage, conversion, and transmission assets to
minimise total system costs in a single linear optimisation problem, which
assumes perfect operational foresight and is solved with \textit{Gurobi
v11}.\cite{gurobi} To manage computational complexity, no pathways with multiple
investment periods are calculated, but overnight scenarios targeting net-zero
CO$_2$ emissions. The capacity expansion is based on technology cost and
efficiency projections for 2040 (see `Data availability'), acknowledging that
much of the required infrastructure must be constructed well before reaching
net-zero emissions. \cref{fig:sensitivity-bars,fig:sensitivity-costs} and
\sfigref{fig:si:infra-e} feature additional scenarios using technology
projections for 2030 and 2050.

Existing hydro-electric power plants\cite{gotzensPerformingEnergy2019} are
included, as well as nuclear power plants built after 1990 or currently under
construction according to Global Energy Monitor's Global Nuclear Plant Tracker
(52~GW total of 106~GW in current
operation).\cite{globalenergymonitorGlobalNuclearPower2024} While
hydroelectricity is assumed to be non-extendable due to geographic constraints,
additional nuclear capacities can be expanded where cost-effective. We assume
the existing nuclear fleet is operated inflexibly and apply country-specific
historical availability factors from 2021 to
2023.\cite{internationalatomicenergyagencyPowerReactorInformation2024}

Temporally, the model is solved with an uninterrupted 4-hourly equivalent
resolution for a single year, using a segmentation clustering approach
implemented in the \textit{tsam} toolbox on all time-varying
data.\cite{hoffmannParetooptimalTemporal2022} While weather variations between
years are not considered for computational reasons, the chosen weather year 2013
is representative in terms of wind and solar availability and heat
demand.\cite{gotskeDesigningSectorcoupledEuropean2024} Some demands are
associated with a time-varying profiles (e.g.~residential/services electricity,
electric vehicles, and heating demand) based on travel patterns or ambient weather
con ditions, while the other exogenous demands are assumed to be time-constant
(e.g.~kerosene, naphtha, methanol, ammonia, and industry electricity). 

Spatially, the model resolves 115 European
regions,\cite{frysztackiStrongEffect2021} covering the European Union, the
United Kingdom, Norway, Switzerland, and the Balkan countries without Malta and
Cyprus. For computational reasons, only electricity, heat, and hydrogen are
modelled at high spatial resolution, while oil, methanol, methane, ammonia, and
carbon dioxide are treated as easily transportable without spatial constraints.
Of the total final energy and non-energy demand
(\sfigref{fig:si:demand_totals}), only some demands are spatially fixed
(\sfigref{fig:si:demands}). These include electricity for residential, industry,
services, and agriculture; heat; electric vehicles; solid biomass for industry;
and naphtha/methanol feedstocks. Hydrogen demand for steel and ammonia
production is also spatially fixed unless these industries can relocate. In this
case, regionally fixed hydrogen demand becomes negligible. Since the model
optimizes the siting and operation of most fuel synthesis units, many demands
are spatially variable (e.g.~electricity demand for electrolysers or hydrogen
demand for methanolisation; see model formulation in \snoteref{sec:si:math}).

A mathematical description of PyPSA-Eur can be found in Section S12 in Neumann
et al.\cite{neumannPotentialRoleHydrogen2023}

\subsection*{Gas and electricity network modelling}
\label{sec:methods-networks}

Networks are considered for electricity, methane, and hydrogen transport.
Existing gas pipelines taken from SciGRID\_gas,\cite{plutaSciGRIDGas2022a} can
be repurposed to hydrogen in addition to new hydrogen
pipelines.\cite{neumannPotentialRoleHydrogen2023} Data on the gas transmission
network is further supplemented by the locations of fossil gas extraction sites
and gas storage facilities based on SciGRID\_gas,\cite{plutaSciGRIDGas2022a} as
well as investment costs and capacities of LNG terminals in operation or under
construction from Global Energy Monitor's Europe Gas
Tracker.\cite{globalenergymonitorEuropeGasTracker2024} Geological potentials for
hydrogen storage are taken from Caglayan et
al.,\cite{caglayanTechnicalPotentialSalt2020} restricting where this low-cost
storage option is available. In modelling gas and hydrogen flows, we incorporate
electricity demands for compression of 1\% and 2\% per 1000km of the transported
energy, respectively.\cite{gasforclimateEuropeanHydrogen2021} Existing
high-voltage grid data is taken from
OpenStreetMap.\cite{xiongModellingHighVoltageGrid2024} For HVDC transmission
lines, we assume 2\% static losses at the substations and additional losses of
3\% per 1000km. The losses of high-voltage AC transmission lines are estimated
using the piecewise linear approximation from Neumann et
al.,\cite{neumannAssessmentsLinear2022} in addition to applying linearised power
flow equations.\cite{horschLinearOptimal2018} Up to a maximum capacity increase
of 30\%
%, we consider dynamic line rating (DLR), leveraging the cooling effect of
wind and low ambient temperatures to exploit existing transmission assets
fully.\cite{glaumLeveragingExisting2023} To approximate $N-1$ resilience,
transmission lines may only be used up to 70\% of their rated dynamic
capacity.\cite{shokrigazafroudiTopologybasedApproximations12022} To prevent
excessive expansion of single connections, power transmission reinforcements
between two regions are limited to 15 GW, while an upper limit of 50.7 GW is
placed on hydrogen pipelines, which corresponds to three 48-inch
pipelines.\cite{gasforclimateEuropeanHydrogen2021}

\subsection*{Wind and solar potentials}
\label{sec:methods-wind-solar}

Renewable potentials and time series for wind and solar electricity generation
are calculated with \textit{atlite},\cite{hofmannAtliteLightweight2021}
considering land eligibility constraints like nature reserves, excluded land use
types, topography, bathymetry, and distance criteria to settlements. Given low
onshore wind expansion in many European countries in recent
years,\cite{ourworldindataInstalledWind2023} a deployment density of 1.5
MW~km$^{-2}$ is assumed for eligible land for onshore wind
expansion.\cite{turkovskaLanduseRequirementsSolar2023a} For reference, this
assumption leads to an onshore wind potential for Germany of 244~GW. The
temporal renewable generation potential for the available area is then assessed
based on reanalysis weather data, ERA5,\cite{ecmwf} and satellite observations
for solar irradiation, SARAH-3,\cite{pfeifrothSurfaceRadiationData2023} in
combination with standard solar panel and wind turbine models provided by
\textit{atlite}.

\subsection*{Biomass potentials}
\label{sec:methods-biomass}

Biomass potentials are restricted to residues from agriculture and forestry, as
well as waste and manure, based on the regional medium potentials specified for
2050 in the JRC-ENSPRESO database.\cite{ruizENSPRESOOpen2019} Continued use of
energy crops or biomass imports are not considered. The finite sustainable
biomass resource can be employed for low-temperature heat provision in
industrial applications, biomass boilers, and CHPs, and (electro-)biofuel
production for use in aviation, shipping, and the chemicals industry.
Additionally, we allow biogas upgrading, including capturing the CO$_2$
contained in biogas, which unlocks all considered uses of regular methane
(\sfigref{fig:si:supply-consumption-options}). The total assumed bioenergy
potentials are 1,372~TWh, which splits into 358~TWh/a for biogas and 1,014~TWh/a
for solid biomass. The total carbon content corresponds to
605~Mt$_{\text{CO}_2}$~a$^{-1}$, which is not fully available as a feedstock for
fuel synthesis or sequestration for negative emissions due to imperfect capture
rates of up to 90\%. Biogenic CO$_2$ can be captured from biogas upgrading,
biomass CHPs and biomass-based low-temperature heat provision in industrial use,
if the added cost of carbon capture is economically viable.

\subsection*{Carbon management}
\label{sec:methods-carbon}

The carbon management features of the model trace the carbon cycles through
various conversion stages: industrial emissions, biomass and gas combustion,
carbon capture in numerous applications, direct air capture, intermediate
storage, electrofuels, recycling, landfill or long-term sequestration. The
overall annual sequestration of CO$_2$ is limited to
200~Mt$_{\text{CO}_2}$~a$^{-1}$, similar to the 250~Mt$_{\text{CO}_2}$~a$^{-1}$
highlighted in the European Commission's carbon management
strategy.\cite{europeancommissionAmbitiousIndustrialCarbon2024} This number
allows for sequestering the industry's unabated fossil emissions (e.g.~in the
cement industry) while minimising reliance on carbon removal technologies. A
carbon dioxide network topology is not co-optimised since CO$_2$ is not
spatially resolved. This means that the location of biogenic or industrial point
sources of CO$_2$ is not a siting factor that this model version considers for
PtX processes, implicitly assuming that the CO$_2$ would be transported there at
low cost. \cite{hofmannDesigningCO22023,hofmannH2CO2Network2024}

\subsection*{Transport sector fuel assumptions}
\label{sec:methods-transport}

While the shipping sector is assumed to use methanol as fuel given its high
technology-readiness level compared to hydrogen or
ammonia,\cite{ieaETPCleanEnergy2024} land-based transport, including heavy-duty
vehicles, is fully electrified in the presented
scenarios.\cite{linkRapidlyDecliningCosts2024} Aviation can use green kerosene
derived from Fischer-Tropsch fuels or methanol, owing to lower technology
readiness levels of fuel cell or battery-electric
aircraft.\cite{ieaETPCleanEnergy2024} Alternative uses for methanol and
Fischer-Tropsch fuels extend beyond transport, including
power-to-methanol,\cite{brownUltralongdurationEnergyStorage2023} diesel for
agriculture machinery and as feedstock for high-value chemicals.

\subsection*{Technical constraints of synthetic fuel production}
\label{sec:methods-efuels}

We consider potential flexibility restrictions in the synthesis processes to
obtain more realistic operational patterns of green electrofuel synthesis
plants. We apply a minimum part load of 20\% for methanolisation and 50\% for
methanation and Fischer-Tropsch
synthesis.\cite{mucciPowerXProcessesBased2023,wentrupDynamicOperationFischerTropsch2022,dieterichPowerliquidSynthesisMethanol2020,mbathaPowermethanolProcessReview2021}
The assumed lower operational flexibility is a potential disadvantage of
Fischer-Tropsch over methanol synthesis, where theses fuels compete. These
`green' options then compete with `blue' and `grey' options, such as steam
methane reforming of fossil gas with or without carbon capture for hydrogen
(\sfigref{fig:si:supply-consumption-options}). Some carriers also feature a
biogenic production route (e.g., methane and oil).

\subsection*{Heating sector modelling and PtX waste heat}
\label{sec:methods-heating}

Heating supply technologies like heat pumps, electric boilers, gas boilers, and
combined heat and power (CHP) plants are endogenously optimised separately for
decentral use and central district heating. District heating shares of demand
are exogenously set to a maximum of 60\% of the total urban heat demand with
sufficiently high population density. Besides the options for long-duration
thermal energy storage, district heating networks can further be supplemented
with waste heat from various power-to-X processes: electrolysis, methanation,
ammonia synthesis, and Fischer-Tropsch fuel synthesis. Because the thermal
discharge from the methanol synthesis is primarily used to distillate the
methanol-water output mix,\cite{brownUltralongdurationEnergyStorage2023} its
waste heat potential is not considered for district heat. Here, we assume a
utilizable share of waste heat of 25\%, considering that within the 115 regions,
only a fraction of fuel synthesis plants might be connected to district heating
systems. In further sensitivity analyses, we explore the effect of no or full
waste heat utilisation.

\subsection*{Backup heat and power options}
\label{sec:methods-backup}

The model includes a variety of options for providing backup power and heating
in periods of low renewable generation and high demand
(\sfigref{fig:si:supply-consumption-options}). Backup power options include
hydrogen, gas and methanol turbines. Backup heat options include gas boilers and
resistive heaters. For combined backup heat and power, we consider biomass,
hydrogen, and gas CHPs. Furthermore, flexible demands like electric vehicles,
heat pumps and fuel synthesis units, as well as batteries and thermal storage in
district heating, can be utilised to reduce the need for backup capacities. 

\subsection*{Industry relocation modelling for steel and ammonia production}
\label{sec:methods-industry}

Unless indicated otherwise, all scenarios also allow the model to relocate the
steel and ammonia industry within Europe endogenously. This allows the best
sites within Europe to compete with outsourced production abroad. While this
captures some of the most energy-intensive industry sectors, other sectors, like
concrete and alumina production, are not considered for relocation.

Without relocation of steel and ammonia production allowed, the production
volumes of primary steel, by direct iron reduction (DRI) and electric arc
furnace (EAF), and ammonia for fertilizers, by Haber-Bosch synthesis, are
spatially fixed. This results in exogenous hydrogen demand per region. Total
production volumes are based on current
levels.\cite{unitedstatesgeologicalsurveyAmmoniaProductionCountry2022,europeancommission.jointresearchcentre.JRCIDEES2021IntegratedDatabase2024}
For the spatial distribution, we use data on the existing integrated steelworks
listed in Global Energy Monitor's Global Steel Plant Tracker
\cite{globalenergymonitorGlobalSteelPlant2024} and manually collected data on
the location and size of ammonia plants in Europe.

With the relocation of steel and ammonia production allowed, the model
endogenously chooses the regional production volumes of primary steel, HBI, and
ammonia, subject to the availability of cheap hydrogen. Thereby, the regional
capacities and operation of Haber-Bosch, DRI, and EAF plants are co-optimised
with the rest of the system, similar to the siting of Fischer-Tropsch or
methanolisation plants. For DRI and EAF, investment costs and specific
requirements for fuels and iron ore are taken from the Steel Sector Transition
Strategy Model (ST-STSM) of the Mission Possible
Partnership.\cite{missionpossiblepartnershipSteelSectorTransition2022,missionpossiblepartnershipMakingNetZeroSteel2022}.
and assume steel can be stored and transported without constraints within
Europe. 

For both cases, we assume a rise in the steel recycling rate from 40\% today to
70\% in our carbon-neutral
scenarios.\cite{materialeconomicsIndustrialTransformation20502019} We assume
that the electric arc furnaces for secondary steel remain, in proportion, at
current locations and do not relocate.

A limitation of the relocation modelling of steel and ammonia production is that
it only considers the cost of energy in the siting of these industries. Other
factors, such as impacts on regional economies and local jobs, integration with
other production processes, or availability of existing infrastructure are not
considered, largely due to a lack of data. The resulting relocation patterns
should therefore be interpreted with caution as they might underestimate total
relocation costs and frictions. we allow domestic relocation nevertheless in
most scenarios, as it would be inconsistent to allow steel and ammonia imports
from abroad while preventing relocation within Europe.

\subsection*{TRACE: Import supply chain modelling}
\label{sec:methods-trace}

The European energy system model is extended with data from the TRACE model
(derivative of v1.1) used in Hampp et al.\cite{hamppImportOptions2023} to assess
the unit costs of different vectors for importing green energy and material to
entry points in Europe from various world regions. For consistency with the
European model, the techno-economic assumptions were aligned, using the same
projections for 2040 (plus 2030 / 2050 in
\cref{fig:sensitivity-bars,fig:sensitivity-costs} and \sfigref{fig:si:infra-e})
and a uniform weighted average cost of capital (WACC) of
7\%.\cite{lonerganImprovingRepresentationCost2023} As possible import vectors,
we consider electricity by transmission lines, hydrogen as gas by pipelines and
as a liquid by ship, methane as a liquid by ship, liquid ammonia, steel and HBI,
methanol and Fischer-Tropsch fuels by ship. Liquid organic hydrogen carriers
(LOHC) are not considered as export vectors due to their lower technology
readiness level (TRL) compared to other
vectors.\cite{irenaGlobalHydrogenTrade2022}

Our selection of 53 potential exporting regions broadly comprises countries with
favourable wind and solar resources and large enough potentials for substantial
exports above 500~TWh~a$^{-1}$ in addition to domestic consumption. We exclude
some countries due to political instability (e.g.~Sudan, Somalia, Yemen), using
a Fragile States Index\cite{thefundforpeaceffpFragileStatesIndex2023} value of
100 as a threshold, or due to severe imposed sanctions (e.g.~Russia, Iran,
Iraq), following the EU Sanctions
Map.\cite{estonianpresidencyofthecounciloftheeuEUSanctionsMap2024} Landlocked
countries without access to seaports or realistic pipeline connections are
excluded. For landlocked regions within pipeline reach, we only exclude
shipborne vectors. Some large countries are split into multiple subregions for
a more differentiated view (e.g.~USA, Argentina, Brazil, and China). The
resulting regions are marked in \cref{fig:options:global}.

% renewable potentials minus domestic demand

To determine the levelised cost of energy for exports, the methodology first
assesses the regional potentials for solar, onshore, and offshore wind energy.
These potentials and time series are calculated using
\textit{atlite},\cite{hofmannAtliteLightweight2021} applying similar land
eligibility constraints as in PyPSA-Eur (but using other datasets with global
coverage) and applying the same wind turbine and solar panel models to
ERA5\cite{ecmwf} weather data for 2013 in eligible regions. Since TRACE
evaluates whole regions without further transmission network resolution, the
renewable potentials and profiles within a region are split into different
resource classes to reduce smoothing effects. We consider 30 classes each for
onshore wind and solar and 10 for offshore wind where applicable. Based on these
calculations, levelised cost of electricity (LCOE) curves can be determined for
each region. A selection of LCOE curves is shown in
\sfigref{fig:si:cost-supply-curves}.

% deducted domestic demand

In the next step, potentials are reduced by the projected future local energy
demand, starting with the lowest LCOE resource classes. With this approach,
domestic consumption is prioritised and supplied by the regions' best renewable
resources, even though we do not model the energy transition in exporting
regions in detail. To create the demand projections, we use the
GEGIS\cite{mattssonAutopilotEnergyModels2021} tool, which utilises machine
learning on historical time series, weather data, and macro-economic factors to
create artificial electricity demand time series based on population and GDP
growth scenarios following the SSP2 scenario of the Shared Socioeconomic
Pathways.\cite{riahiSharedSocioeconomicPathways2017} From these time series we
take the annual total and increase it by a factor of two to account for further
electrification of other sectors, which the GEGIS tool does not consider.

% conversion pathways and scale

The remaining wind and solar electricity supply can then be used to produce the
specific energy or material vector according to the flow chart of conversion
pathways shown in \sfigref{fig:si:import-esc-scheme}. Considered technologies
include water electrolysis for \ce{H2}, direct air capture (DAC) for \ce{CO2},
synthesis of methane, methanol, ammonia or Fischer-Tropsch fuels from \ce{H2}
with \ce{CO2} or \ce{N2}, and \ce{H2} direct iron reduction (DRI) for sponge
iron with subsequent processing to green steel in electric arc furnaces (EAF)
from iron ore priced at
97.7~\euro{}~t$^{-1}$.\cite{missionpossiblepartnershipSteelSectorTransition2022}
Other CO$_2$ sources than DAC are not considered in the exporting regions.
Furthermore, while batteries and hydrogen storage in steel tanks are considered,
underground hydrogen storage is excluded due to the uncertain availability of
salt caverns in many potential exporting
regions.\cite{hevinUndergroundStorageHydrogen2019,hydrogentcp-task42UndergroundHydrogenStorage2023}
We also assume that the energy supply chains dedicated to exports will be
islanded from the rest of the local energy system, i.e.~that curtailed
electricity or waste heat could not be used locally.

For each vector, an annual reference export demand of 500~TWh$_\text{LHV}$ or
100~Mt of steel and HBI is assumed, mirroring large-scale energy and material
infrastructures and export volumes, corresponding to approximately 40\% of
current European LNG
imports\cite{instituteforenergyeconomicsandfinancialanalysisEuropeanLNG2023} and
66\% of European steel
production.\cite{eurofer-theeuropeansteelassociationEuropeanSteel2023} Transport
distances are calculated between the exporting regions and the twelve
representative European import locations using the \textit{searoute} Python
tool\cite{haliliSeaRoutePython2024} for shipborne vectors or crow-fly distances
for pipeline or HVDC connections, and modified by a mode-specific detour factor.
The chosen representative import locations are based on large ports and LNG
terminals in the United Kingdom, the Netherlands, Poland, Greece, Italy, Spain,
and Portugal, as well as pipeline entry points in Slovakia, Greece, Italy, and
Spain. All energy supply chains are assumed to consume their energy vector as
fuel for transport to Europe, except for HBI and steel, which use externally
bought green methanol as shipping fuel. The capital costs of the ships and
pipelines are also included, following the metholodogy of Hampp et
al.~\cite{hamppImportOptions2023}.

%%% capacity expansion %%%

For each combination of carrier, exporter, and importer, a linear capacity
expansion optimisation is performed to determine cost-optimal investments and
the operation of generation, conversion, storage, and transport capacities for
all intermediary products to deliver 500~TWh~a$^{-1}$ (or 100~Mt~a$^{-1}$ for
materials) of the final carrier to Europe. Dividing the total annual system
costs by the targeted annual export volume yields the levelised cost of energy
or material as seen by the European entry point. To match the multi-hourly
resolution used for the European model, the TRACE model was configured to use a
3-hourly resolution for 2013, resulting in similar balancing requirements.
Considering the reference export volume of 500~TWh~a$^{-1}$ (or 100~Mt~a$^{-1}$
for materials), the resulting levelised cost curves of imports for different
import vectors and exporting regions are presented for the respective
lowest-cost entry point to Europe in
\sfigref{fig:si:isc-h2,fig:si:isc-ch4-nh3,fig:si:isc-meoh-ft,fig:si:isc-hbi-St}.
The curves show the varying cost composition of the country-carrier pairs. In
this step, each import vector combination of carrier, exporter, and importer is
optimised separately. Further constraints, like constraints on total export
volumes per country, are imposed in the coupling to the European model.

A mathematical description of TRACE can be found in Section S3 in Hampp et
al.\cite{hamppImportOptions2023}

\subsection*{Coupling of import options to European model}
\label{sec:methods-coupling}

The resulting levelised unit cost for each combination of carrier, exporter, and
reference importer is then used as an exogenous input to the European model. For
each candidate entry point in the 115 European model regions, we match the
closest reference import location from TRACE and add the corresponding import
cost curve as a supply option
(\sfigref{fig:si:isc-h2,fig:si:isc-ch4-nh3,fig:si:isc-meoh-ft,fig:si:isc-hbi-St}).  
Moreover, we limit energy exports from any one exporting region to Europe for
the sum of all carriers to 500~TWh~a$^{-1}$. This is to both prevent a single
country from dominating the import mix and be consistent with the target export
volume assumed in TRACE. Beyond that, the decision about the origin,
destination, vector, volume, and timing of imports is largely endogenous to
PyPSA-Eur.

%%% entry points %%%

However, imports may be further restricted by the expansion of domestic import
infrastructure. For each vector, we identify locations where the respective
carrier may enter the European energy system by considering where LNG terminals
and cross-continental pipelines are located (\cref{fig:options:europe}). For
hydrogen imports by pipeline, imports must be near-constant, varying between
90-100\% of peak imports, aligning with the high pipeline utilisation rates
observed in the TRACE model. For methane imports by ship, existing LNG terminals
reported in Global Energy Monitor's Europe Gas
Tracker\cite{globalenergymonitorEuropeGasTracker2024} can be used. For hydrogen
by ship, new terminals can be built in regions where LNG terminals exist. To
ensure regional diversity in potential gas and hydrogen imports and avoid
vulnerable singular import locations, we allow the expansion beyond the reported
capacities only up to a factor of 2.5, taking the median value of reported
investment costs for LNG terminals.\cite{GlobalGas2022} A premium of 20\% is
added for hydrogen import terminals due to the lack of practical experience with
them. For electricity, the capacity and operational patterns of the HVDC links
can be endogenously optimised. Imports for carbonaceous fuels, ammonia, HBI, and
steel are not spatially allocated to specific ports, given their low transport
costs relative to value. Port capacities are assumed unconstrained since these
commodities, particularly carbonaceous fuels, are comparable to the large fossil
oil volumes currently handled at European ports.

%%% reconversion %%%

Further conversion of imported fuels is also possible once they have arrived in
Europe, e.g.~hydrogen could be used to synthesise carbon-based fuels, ammonia
could be cracked to hydrogen, methane could be reformed to hydrogen, and methane
or methanol could be combusted for power generation. However, conversion losses
can make it less attractive economically to use a high-value hydrogen derivative
merely as a transport and storage vessel only to reconvert it back to hydrogen
or electricity.

%%% special handling of electricity imports %%%

The supply chain of electricity imports is endogenously optimised with the rest
of the European system rather than using a constant levelised cost of
electricity for each export region. This is because, owing to the greater
challenge of storing electricity, the hourly variability of wind and solar
electricity leads to higher price variability than hydrogen and its derivatives,
and the intake needs to be more closely coordinated with the European power
grid. The endogenous optimisation comprises wind and solar capacities, batteries
and hydrogen storage in steel tanks, and the size and operation of HVDC link
connections to Europe based on the renewable capacity factor time series as
illustrated in \cref{fig:options:europe}. Europe's connection options with
exporting regions are confined to the 4\% nearest regions, with additional
ultra-long distance connection options to Ireland, Cornwall, and Brittany
following the vision of the Xlinks project between Morocco and the United
Kingdom.\cite{xlinksMoroccoUKPowerProject2023} Connections through Russia or
Belarus are excluded. In addition to excluded entry points, some connections
from Central Asia are affected by additional detours beyond the regularly
applied detour factor of 125\% of the as-the-crow-flies distance. Similar to
intra-European HVDC transmission, a 3\% loss per 1000km and a 2\% converter
station loss are assumed.

\vspace{1em}

Finally, we note that all mass-energy conversion is based on the lower heating
value (LHV). To present energy and material imports in a common unit, the
embodied energy in steel is approximated with the 2.1 kWh~kg$^{-1}$ released in
iron oxide reduction, i.e.~energy released by
combustion.\cite{kuhnIronRecyclable2022} All currency values are given in
\euro{}$_{2020}$.
